---
title: "Demultiplex hashtags in SLN208_D1 (Lane 6)"
output: html_notebook
---

Zoë Steier

Required inputs: 
  outs_dir (aggregated 10x output)

1. Create metadata
2. Demultiplexing hashtags with Seurat
3. Save outputs for totalVI

# Preprocessing and creating metadata

## Load packages and data
```{r Load required packages}
library(Seurat)
library(tidyverse)
library(Matrix)
```


```{r Load feature-barcode matrix from cellranger output}
# Matrix contains genes and proteins
outs_dir <- "/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/cellranger_out/6_B6_spl_LN_r4_206/outs/"

matrix_dir <- str_c(outs_dir, "filtered_feature_bc_matrix/")
barcode.path <- paste0(matrix_dir, "barcodes.tsv.gz")
features.path <- paste0(matrix_dir, "features.tsv.gz")
matrix.path <- paste0(matrix_dir, "matrix.mtx.gz")
mat <- readMM(file = matrix.path) # Sparse dgTMatrix (genes x cells)
feature.names <- read.delim(features.path, # Features are the genes
                           header = FALSE,
                           stringsAsFactors = FALSE)
barcode.names <- read.delim(barcode.path, # Barcodes are the cells
                           header = FALSE,
                           stringsAsFactors = FALSE)
```

## Preprocess data into matrices

```{r Separate RNA, ADT, and HTO into separate matrices}
# Currently they are in a single matrix. Genes are first, followed by proteins.
num_proteins <- sum(grepl("Antibody Capture", feature.names$V3)) 
num_features <- dim(feature.names)[1]

# Make protein matrix
protein_features <- feature.names[(num_features - num_proteins + 1):num_features,]
protein_mat <- mat[(num_features - num_proteins + 1):num_features,]
colnames(protein_mat) <- barcode.names[,]
row.names(protein_mat) <- protein_features[,"V2"]

# Separatae ADT and HTO into two matrices
hto_start <- grep("HTO", row.names(protein_mat))[1] # find first protein name that is an HTO 
ADT_mat <- protein_mat[1:(hto_start - 1), ]
HTO_mat <- protein_mat[hto_start:num_proteins,] # HTOs are at the end

# Make rna matrix
gene_features <- feature.names[1:(num_features - num_proteins),]
mat <- mat[1:(num_features - num_proteins),]
```


```{r Convert ensembl IDs into gene symbols}
# Get gene names from the feature.names. Not all gene names are unique, so take the summed expression.
# feature.names$V2 are gene names, $V1 are ensembl ids

# Create expression matrix with unique gene symbols

#' exp is expression matrix
#'
#' gene_symbol is a character vector, length is rows of exp
#'   contains symbol for gene for each row
#' ensemble_id is a character vector, length is rows of exp
#'   contains unique ensemble id for each row
agg_sparse_mat <- function(exp, gene_symbol, ensemble_id){
  #  (gene_symbol x gene_id) x (genes_id x cells) = (gene_symbol x cells)
  unique_symbols <- sort(unique(gene_symbol))
  ens_id <- seq(length(ensemble_id))
  unique_id <- match(gene_symbol, unique_symbols) 
  aggMat <- sparseMatrix(i = unique_id, j = ens_id,
                         dims = c(length(unique_symbols), length(ensemble_id)),
                         dimnames = list(unique_symbols, ensemble_id))
  exp_sym <- aggMat %*% exp 
  return(exp_sym)
}

# Apply the function to generate a matrix of unique gene symbol row names (sparse format)
exp_sym <- agg_sparse_mat(exp = mat, gene_symbol = gene_features$V2, ensemble_id = gene_features$V1)
colnames(exp_sym) <- barcode.names[,] # add column names as cell barcodes
```


```{r Dimensions of feature-barcode matrix}
dim(mat)
dim(exp_sym)
dim(ADT_mat)
dim(HTO_mat)
```

# Make Seurat object and demultiplex

## Make seurat object

```{r Make Seurat object and add RNA}
# Setup Seurat object starting with RNA
seurat_obj <- CreateSeuratObject(counts = exp_sym)

# Normalize RNA data with log normalization
seurat_obj <- NormalizeData(seurat_obj)

# Find and scale variable features
seurat_obj <- FindVariableFeatures(seurat_obj, assay = "RNA", selection.method = "mean.var.plot")

# Scale RNA
seurat_obj <- ScaleData(seurat_obj, features = VariableFeatures(seurat_obj))
```

```{r Add protein data}
# Add protein data as a new assay independent from RNA
seurat_obj[["ADT"]] <- CreateAssayObject(counts = ADT_mat) # Feature names can't have "_", replaces with "-"

# Normalize ADT data, here we use centered log-ratio (CLR) transformation
seurat_obj <- NormalizeData(seurat_obj, assay = "ADT", normalization.method = "CLR")

# Scale ADT data
seurat_obj <- ScaleData(seurat_obj, assay = "ADT")

# See https://github.com/satijalab/seurat/blob/c9f2660ee6b88189acc3ff4aa6088a46a2558dd3/R/preprocessing.R for CLR method. margin = 1
```


```{r Add hashatag data}
seurat_obj[["HTO"]] <- CreateAssayObject(counts = HTO_mat)

# Normalize HTO data, here we use centered log-ratio (CLR) transformation
seurat_obj <- NormalizeData(seurat_obj, assay = "HTO", normalization.method = "CLR")

```

## Demultiplex

```{r Demultiplex hashatgs with kmeans}
seurat_obj <- HTODemux(seurat_obj, assay = "HTO", positive.quantile = 0.995, kfunc = "kmeans") # default is clara
table(seurat_obj$HTO_classification.global)
Idents(seurat_obj) <- "hash.ID"
FeatureScatter(seurat_obj, feature1 = "HTO-B6-LN-r4-206-A0301", feature2 = "HTO-B6-spl-r4-206-A0302")
```

Parameters: kmeans, 0.995 (quantile above negative cells in NB distribution)
Cutoff for HTO-B6-LN-r4-206-A0301 : 10 reads
Cutoff for HTO-B6-spl-r4-206-A0302 : 8 reads

 Doublet Negative  Singlet 
     552      141    10084 
     

```{r Plot HTO results}
Idents(seurat_obj) <- "hash.ID"
RidgePlot(seurat_obj, assay = "HTO", features = rownames(seurat_obj[["HTO"]])[1:2], ncol = 2)
```


```{r Heatmap of HTO results}
HTOHeatmap(seurat_obj, assay = "HTO")

```

```{r Violin plot of HTO results}
Idents(seurat_obj) <- "HTO_classification.global"
VlnPlot(seurat_obj, features = "nCount_RNA", pt.size = 0.1, log = TRUE)
```

# Save output

```{r Save hashtag labels}
#### Note: hashtag order is the opposite of spleenLN111
# 111: spleen is hash1, LN is hash2
# 206: spleen is hash2, LN is hash1

# Add tissue labels and save hash metadata
table(seurat_obj@meta.data$hash.ID)

tissue <- tibble(Cell_BC = row.names(seurat_obj@meta.data), hash.ID = seurat_obj@meta.data$hash.ID) %>%
  mutate(Hash_ID = str_replace(str_replace(hash.ID, "HTO-B6-LN-r4-206-A0301", "Lymph_Node"), "HTO-B6-spl-r4-206-A0302", "Spleen")) %>%
  select(-hash.ID)

table(tissue$Hash_ID)

# Save metadata
write_csv(tissue, "spleenLN206_day1/hashtag_labels.csv")

```

```{r Save seurat object}
# Add hashtag labels to seurat object
seurat_obj <- AddMetaData(seurat_obj, tissue$Hash_ID, col.name = "Hash_label")

saveRDS(seurat_obj, "spleenLN206_day1/seurat_obj.rds")

```